---
title: "simmr: quick start guide"
author: "Andrew Parnell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick_start}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

## Step 1: install `simmr`

Use:
```{r, eval = FALSE}
install.packages('simmr')
```

then
```{r, message=FALSE}
library(simmr)
library(dplyr)
```

## Step 2: load in the data

Some geese isotope data is included with this package. Find where it is with:
```{r, eval = FALSE}
system.file("extdata", "geese_data_small.xls", package = "simmr")
```

Load into R with:
```{r, echo = FALSE}
if (!requireNamespace("readxl", quietly = TRUE)) {
    stop("readxl needed for this vignette to work. Please install it.",
      call. = FALSE)
}
```
```{r}
library(readxl)
path = system.file("extdata", "geese_data_small.xls", package = "simmr")
```

Now load in the data
```{r}
geese_data = lapply(excel_sheets(path), read_excel, path = path)
```

If you want to see what the original Excel sheet looks like you can run `system(paste('open',path))`.

We can now separate out the data into parts 
```{r}
geese_data = list(mixtures = geese_data[[1]],
                  tracer_names = c('d13C_Pl', 'd15N_Pl'),
                  source_names = geese_data[[2]][,1],
                  source_means = geese_data[[2]][,2:3],
                  source_sds = geese_data[[2]][,4:5],
                  correction_means = geese_data[[3]][,2:3],
                  correction_sds = geese_data[[3]][,4:5],
                  concentration_means = geese_data[[4]][,2:3])
```

## Step 3: load the data into `simmr` and create iso-space plot

```{r, fig.width = 7, fig.height = 5}
simmr_geese = geese_data %>% simmr_load
```

## Step 4: run through `simmr` and check convergence

```{r, results = 'hide', message = FALSE}
geese_simmr_out = simmr_geese %>% simmr_mcmc
```
```{r}
geese_simmr_out %>% summary(type = 'diagnostics')
```
All convergence diagnostics should be close to 1

Check that the model fitted well:
```{r, results = 'hide', message = FALSE, fig.width = 7, fig.height = 5}
geese_simmr_out %>% posterior_predictive
```

Half the points should lie in the error bars and the other half outside if it's well-calibrated.

## Step 6: look at the output

Look at the influence of the prior:
```{r, fig.width = 7, fig.height = 5}
geese_simmr_out %>% prior_viz
```

Look at the histogram of the dietary proportions:
```{r, fig.width = 7, fig.height = 5}
geese_simmr_out %>% plot(type = 'histogram')
```

For the many more options available to run and analyse output, see the main vignette via `vignette('simmr')`
